Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-10-18 manuscript.RMD
readme updates

**Purpose:** Muscle hypertrophy, induced by resistance training, is determined by ribosomal biogenesis. High-glucose treatment may augment the transcription of ribosomal RNA, as evident from in vitro studies. However, its effect on resistance training-induced adaptations remains unexplored.

**Methods:** Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 23.7 (1.8)/24.6 (4.8)), participated in a within-participant trial with unilateral resistance training with glucose- and placebo supplementation on alternate days over six sessions per condition. Participants remained fasted overnight, receiving glucose/placebo and protein supplementation only before and after RT sessions (3 sets of 10 repetitions maximum unilateral leg press and knee extension). Micro biopsies were sampled from m. vastus lateralis before and after the intervention.

**Results:** Between conditions, there were no significant differences in increases of total RNA and ribosomal RNA (mean difference `r total_rna_stats`, 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`), or levels of cellular myelocytomatosis oncogene, upstream binding factor and ribosomal protein S6 (*p* = `r min(p_west)`-`r max(p_west)`). There was a linear relationship between upstream binding factor and total RNA, where an increase of 1 standard deviance unit of upstream binding factor equated to a $\sim$ `r corr_incr`% increase in total RNA (*p* = `r corrp`).

**Conclusion:** Glucose supplementation before and after RT does not augment RT-induced ribosome biogenesis after two weeks of heavy RT in moderately trained young adults.

@@

Data was collected at the sports science lab at the Inland Norway University of Applied Sciences, with all analyses except for serum blood samples performed at the sports science lab. Serum blood samples were performed at Innlandet Hospital Trust. S.E., H.H. and D.H. conceived and designed the study; K.L., S.C.M., D.H. and H.H. performed experiments; K.L. and D.H. interpreted the results; K.L. and D.H. prepared figures; K.L. drafted the manuscript; D.H., H.H. and S.E. edited and revised the manuscript; All authors approved the final version of the manuscript.

### Acknowledgements
Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak and Thea Bøhn Nyløkken for their dedication and hard work in the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

**Abbreviations**
```{r abbreviation table}
### Abbreviations table

@@

abb.tab
```

\pagebreak
## Introduction

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. While there certainly may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], there are also recent studies showing that epigenetic factors such as ribosomal DNA (rDNA) availability may also impact the ability to mount an anabolic response to RT [@figueiredo_genetic_2021]. Indeed, the ability to synthesise ribosomes following a period of RT is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020]. For instance, increasing training volume generally induces greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017]. However, as evident from Hammarström et al. (2020), not all participants experienced increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine effectively optimise RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains equivocal if other nutritional adjuvants such as glucose can increase the efficacy of RT [@tezze_anabolic_2023]. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target ribosomal protein S6 kinase beta-1 (S6K1) serves as a mediator for insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures [@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to test the hypothesis that glucose supplementation before and after RT increases markers of ribosomal abundance following five RT sessions. Secondly, we aimed to describe the association between UBF and total RNA abundance in human skeletal muscle. There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency concerning long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucosum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and -c-peptide levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

Two muscle biopsy aliquots were used for total RNA extraction per leg, resulting in a total of eight RNA samples per participant. Total RNA was extracted using TRIzol with muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). To enable analysis of target gene expression per unit tissue weight [@ellefsen_improved_2008; @ellefsen_reliable_2014], an exogenous RNA control (Lambda, λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount to each sample (0.04 ng ml^-1^ of TRIzol reagent). For assessment of RNA content and purity, RNA was eluted in TE-buffer (1:2) and assessed via spectrophotometry. All samples had a 260 nm to 280 nm ratio > 1.9. The RNA stock was stored at -80°C until further analyses. Before quantitative analyses of total RNA, samples with known loss of RNA during extraction (n = `r technical`) or a deviation from the observed RNA to muscle tissue weight relationship larger than 3 $\times$ residual SD while accounting for training status (n = `r observed`) were removed from the data set. Total RNA was normalised to wet muscle weight and log transformed before statistical analyses.

@@

Total protein was extracted using the Minute Total Protein Extraction Kit for Muscles (Invent Biotechnology), according to the manufacturer’s protocol, optimised for our lab. Wet muscle was freeze-dried for 24 hours and dissected before extraction. The tissue was homogenised with a plastic rod in 80 mg protein extraction powder (Invent Biotechnology) and 100 ul ice-cold cell lysis buffer (Denaturing Buffer, Invent Biotechnology), and centrifuged at 19 000 g for 1 minute. The supernatant was divided into aliquots to run samples in duplicates, and total protein concentrations were determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway). The protein samples were diluted to a concentration of 1.5 µg µl^−1^ with lysis buffer and 4X Laemmli sample buffer (Bio-Rad Laboratories AB, Oslo, Norway) containing 2-mercaptoethanol. All protein samples were incubated at 95°C and stored at -20°C until further analysis. The protein samples (20.25 µg total protein) were separated at 250V on 4-20% Tris-Glycine gels (Bio-Rad Laboratories) for 50 minutes and then transferred to PVDF membranes with wet transfer at 300 mA for 3 hours. Both gel electrophoresis and protein transfer were performed at 4°C. Following the wet transfer, membranes were stained with a reversible total protein stain (Thermo Fisher Scientific) and then blocked for 1 hour at room temp with a blocking buffer of Tris-buffered saline (TBS; 20 mM Tris, 150 mM NaCl) with 5% non-fat dry milk and 0.1% Tween-20. Primary and secondary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA): UBF, UBF F-9, sc-13125; rpS6, Ribosomal protein S6 C-8, sc-74459; and Thermo Fisher Scientific (Oslo, Norway): c-Myc, 9E10; goat anti-mouse (for c-Myc), goat anti-mouse IgG1 (y1) horseradish peroxidase conjugate; and anti-mouse (anti-mouse IgG1 horseradish peroxidase conjugate). Antibodies were diluted in blocking buffer to concentrations corresponding to 1:25 000 (UBF, rpS6) and 1:5000 (c-Myc).

@@

A priori power calculations showed that 20 participants would grant a statistical power of 80% (alpha = 0.05), accounting for an expected dropout of 20%. This power calculation was based on an assumption that the effects of glucose ingestion on ribosome biogenesis (measured as within-participant differences in total RNA) may equate to the effects of increasing RT volume from low to moderate [@hammarstrom_benefits_2020]. Total RNA, protein and qPCR data were analyzed by mixed-effects models with fixed effects included as *supplement $\times$ time*. To decrease the risk of Type I errors, random effects were selected from step-wise elimination of terms from the most complex structure (random slopes for time and supplement and their interaction) to less complex. The most complex random effects structure that converged was chosen as the final model[@matuschek_balancing_2017]. Plasma glucose, c-peptide, training volume, and the strength index were analyzed by multiple time-point log-fold change score comparisons, using a mixed-effects model with baseline values, time and supplement, and the interaction between time and supplement as fixed effects. These data only supported a random intercept per participant. The linear mixed-effects models were fitted with the lmer function from the lme4 package using the lmerTest package to procure p-values (Satterthwaite's method for approximating degrees of freedom)[@bates_fitting_2014], written for R[@Rcore]. Log-transformed values were expressed as fold changes in visualisations. Descriptive data are presented as mean and (SD). Inferential statistics are presented as means with 95% confidence intervals and *p*-values unless otherwise stated. *p* < 0.05 was considered statistically significant.

@@

There was no difference in mean change of total training session volume between exercising with placebo or glucose, with an increase of 18 and 16% from session 1 to session 6, respectively (*p* > 0.05). There were no differences in mean macronutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* > 0.05 for all).

@@

Glucose ingestion before and after RT led to significant increases in plasma glucose levels by 38% immediately before RT (Figure 1B, 0 min), by 31% during RT (Figure 1B, 15 min) and by 32% immediately after RT (Figure 1B, 30 min; all : *p* < 0.001), compared to placebo. Two hours after the session, RT with glucose measured 8% lower plasma glucose levels compared to RT with placebo (Figure 1B, 270min: *p* = `r p.270`). Similarly, levels of c-peptide increased with glucose ingestion over placebo, by 95% immediately before (Figure 1C, 0 min) and 87% after RT (Figure 1C, 30 min; both *p* < 0.001).

@@

After 5 RT sessions (Figure 1D, Post 5RT), there was a significant difference in strength between glucose and placebo,  where placebo decreased strength by 7% more than glucose (Figure 1D, *p* = `r pdiff.c4`). At the other time points, there were no significant differences in mean change between exercising with glucose and placebo (Figure 1D, Post 2RT: *p* = `r pdiff.c2`, Post 4RT: *p* = `r pdiff.c3`, 30min post 6RT: *p* = `r pdiff.c5`, 2h post 6RT: *p* = `r pdiff.c6`, 23h post 6RT: *p* = `r pdiff.c7`). Both RT with glucose and placebo led to significantly reduced strength after the fifth session compared to baseline, by 11 and 18% respectively (Figure 1D, Post 5RT: *p* = 0.000). Comparisons of the acute data gathered from after five sessions until and including 23 hours after the sixth session showed a significant mean increase in strength of 5-9% from RT with glucose and placebo 30 minutes after the sixth RT session (Figure 1D, 30min post 6RT: *p* = `r apc2`) and two hours after the sixth RT session (Figure 1D, 2h post 6RT: *p* = `r apc3`). Twenty-three hours after the last (sixth) RT session, strength was unchanged compared to after the fifth RT session (Figure 1D, 23h post 6RT: *p* = `r apc4`).

@@

From baseline to after the intervention total RNA and rRNA per unit muscle weight increased on average by ~20-27% and ~25-57%, respectively (Figure 2). However, RT with glucose did not induce increased accumulation of total RNA (Figure 2A, mean difference `r total_rna_stats`) or rRNA  (Figure 2B; 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`) compared to RT with placebo.

@@

Complete data sets and scripts are downloadable here; https://github.com/Kristianlian/ribose-paper

@@

Extended methods can be found here; https://github.com/Kristianlian/ribose-paper

\pagebreak
## References

<div id="refs"></div>

>>>>

**Purpose:** Resistance training-induced skeletal muscle hypertrophy seems to depend on ribosome biogenesis and content. High glucose treatment may augment ribosomal biogenesis, as evident from *in vitro* studies, and may thus act to potentiate resistance training-induced adaptations *in vivo*. However, this remains largely unexplored.

**Methods:** Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 24.1 (3.3)), participated in a within-participant crossover trial, conducting unilateral resistance training (leg press and knee extension, 3 sets of 10 repetitions maximum) with pre- and post-exercise ingestion of either glucose (3x30g, 90g total) or placebo supplement (Stevia rebaudiana extract, 3x0.3g, 0.9g total), together with protein (2x25g, 50g total), on alternating days for twelve days; six sessions per condition, resistance training was performed in an otherwise fasted state, as morning sessions. Micro biopsies were sampled from m. vastus lateralis before and after the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4/5, 8/9) and after the intervention.

**Results:** Glucose ingestion did not have beneficial effects on resistance training-induced increases of ribosomal content (mean difference 7.6% [-7.2, 24.9], *p* = 0.34; ribosomal RNA, 47S/18S/28S/5.8S/5S, range 7.6%-37.9%, *p* = 0.40-0.98) or levels of relevant transcriptional or translational regulators (c-MYK/UBF/rpS6, *p* = 0.094-0.292). Of note, both baseline and trained state data of total RNA showed a linear relationship with UBF; a ∼14% increase in total RNA corresponded to 1 standard deviation unit increase in UBF (*p* = 0.003).

**Conclusion:** Glucose ingestion before and after resistance training did not augment ribosomal RNA accumulation during twelve days of heavy-load resistance training in moderately trained young adults.

@@

Data was collected at the sports science lab at the Inland Norway University of Applied Sciences, with all analyses performed on-site, except serum blood analyses, which were performed at Innlandet Hospital Trust. S.E., H.H. and D.H. conceived and designed the study; K.L., S.C.M., D.H. and H.H. performed experiments; K.L. and D.H. interpreted the results; K.L. and D.H. prepared figures; K.L. drafted the manuscript; D.H., H.H. and S.E. edited and revised the manuscript; All authors approved the final version of the manuscript.

### Acknowledgements

Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak and Thea Bøhn Nyløkken for their dedication and hard work during the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

**Abbreviations**

```{r abbreviation table}
### Abbreviations table

@@

abb.tab
```
:::

\pagebreak

## Introduction

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. There may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], however, the internal physiological milieu seems favourably shaped for muscle growth by factors such as different types of training stress and nutrients [@figueiredo_genetic_2021; @sebata_control_2018; @thalacker-mercer_cluster_2013]. Indeed, training stress and nutrients impact the ability to synthesise ribosomes, which in turn is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020; @kusnadi_regulation_2015; @sebata_control_2018]. For instance, increasing training volume generally induces greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020; @krieger_single_2009; @schoenfeld_strength_2017]. Still, as evident from Hammarström et al. (2020), not all participants experience increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine are often used to optimise RT adaptations [@cermak_protein_2012; @lanhers_creatine_2015; @lanhers_creatine_2017; @morton_systematic_2018]. However, it remains equivocal if other nutritional adjuvants such as glucose can increase the efficacy of RT [@tezze_anabolic_2023]. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015; @sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015; @moss_housekeeper_2007; @sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019; @hammarstrom_benefits_2020; @sebata_control_2018; @stec_ribosome_2016; @von_walden_ribosome_2019]. In addition, insulin may have anabolic effects with elevated amino acids, and has a clear role in the reduction of muscle protein breakdown independent of amino acid availability [@abdulla_role_2016; @hillier_physiological_2000]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Ribosomal biogenesis and content seem to be a prerequisite for skeletal muscle growth, and transcription of ribosomal ribonucleic acid (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995]. Multiple signalling pathways converge to regulate rRNA transcription, including c-Myc and the mammalian target of rapamycin complex 1 (mTORC1) signal-transduction pathway [@kusnadi_regulation_2015; @von_walden_ribosome_2019; @sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. The general transcription factor c-Myc increases ribosomal biogenesis directly through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011; @mori_c-myc_2021; @west_acute_2016]. Indeed, UBF phosphorylation, which is required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, seems to be increased by high-glucose treatment in a mTORC1-dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011]. Independent of UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures [@zhai_human_2012]. The mTORC1 pathway receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009] and contributes to ribosome biogenesis through the forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription and through the regulation of ribosomal protein translation [@figueiredo_regulation_2019; @von_walden_ribosome_2019]. Moreover, mTORC1 and its downstream target ribosomal protein S6 kinase beta-1 (S6K1) serves as direct mediators of insulin signalling in skeletal muscle [@hillier_physiological_2000]. Together these observations indicate a potential role of glucose in positively affecting human skeletal muscle rRNA synthesis, through potentiating transcription or translation of ribosomal RNA and increasing the translational capacity of skeletal muscle cells [@sebata_control_2018; @zhai_human_2012; @hillier_physiological_2000; @hoppe_amp-activated_2009].

There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency concerning long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011; @von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

Therefore, the purpose of this investigation was to test the hypothesis that glucose supplementation before and after five RT sessions over 12 days will potentiate RT-associated accumulation of markers of ribosomal abundance following five RT sessions. Secondly, we aimed to describe the association between changes in total RNA abundance and UBF in human skeletal muscle.

@@

The study was designed as a 12-day double-blinded placebo-controlled simultaneous crossover trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucoseum monohydricum vs. \~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and serum c-peptide levels). Participants were asked to avoid resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

Two muscle biopsy aliquots were used for total RNA extraction per leg, resulting in a total of eight RNA samples per participant. Total RNA was extracted using TRIzol with muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (\~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). To enable analysis of target gene expression per unit tissue weight [@ellefsen_improved_2008; @ellefsen_reliable_2014], an exogenous RNA control (Lambda, λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount to each sample (0.04 ng ml^-1^ of TRIzol reagent). For assessment of RNA content and purity, RNA was eluted in TE buffer (1:2) and assessed via spectrophotometry. All samples had a 260 nm to 280 nm ratio \> 1.9. The RNA stock was stored at -80°C until further analyses. Before quantitative analyses of total RNA, samples with known loss of RNA during extraction (n = `r technical`) or a deviation from the observed RNA to muscle tissue weight relationship larger than 3 $\times$ residual SD while accounting for training status (n = `r observed`) were removed from the data set. Total RNA was normalised to wet muscle weight and log transformed before statistical analyses.

@@

Total protein was extracted using the Minute Total Protein Extraction Kit for Muscles (Invent Biotechnology), according to the manufacturer's protocol, optimised for our lab. Wet muscle was freeze-dried for 24 hours and dissected before extraction. The tissue was homogenised with a plastic rod in 80 mg protein extraction powder (Invent Biotechnology) and 100 ul ice-cold cell lysis buffer (Denaturing Buffer, Invent Biotechnology), and centrifuged at 19 000 g for 1 minute. The supernatant was divided into aliquots to run samples in duplicates, and total protein concentrations were determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway). The protein samples were diluted to a concentration of 1.5 µg µl^−1^ with lysis buffer and 4X Laemmli sample buffer (Bio-Rad Laboratories AB, Oslo, Norway) containing 2-mercaptoethanol. All protein samples were incubated at 95°C and stored at -20°C until further analysis. The protein samples (20.25 µg total protein) were separated at 250V on 4-20% Tris-Glycine gels (Bio-Rad Laboratories) for 50 minutes and then transferred to PVDF membranes with wet transfer at 300 mA for 3 hours. Both gel electrophoresis and protein transfer were performed at 4°C. Following the wet transfer, membranes were stained with a reversible total protein stain (Thermo Fisher Scientific) and then blocked for 1 hour at room temp with a blocking buffer of Tris-buffered saline (TBS; 20 mM Tris, 150 mM NaCl) with 5% non-fat dry milk and 0.1% Tween-20. Primary and secondary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA): UBF, UBF F-9, sc-13125; rpS6, Ribosomal protein S6 C-8, sc-74459; and Thermo Fisher Scientific (Oslo, Norway): c-Myc, 9E10; goat anti-mouse (for c-Myc), goat anti-mouse IgG1 (y1) horseradish peroxidase conjugate; and anti-mouse (anti-mouse IgG1 horseradish peroxidase conjugate). Antibodies were diluted in blocking buffer to concentrations corresponding to 1:25 000 (UBF, rpS6) and 1:5000 (c-Myc).

@@

A priori power calculations showed that 20 participants would grant a statistical power of 80% (alpha = 0.05), accounting for an expected dropout of 20%. This power calculation was based on an assumption that the effects of glucose ingestion on total RNA accumulation and rRNA expression may equate to the effects of increasing RT volume from low to moderate [@hammarstrom_benefits_2020]. Total RNA, protein and qPCR data were analyzed by mixed-effects models with fixed effects included as *supplement* $\times$ time. To decrease the risk of Type I errors, random effects were selected from step-wise elimination of terms from the most complex structure (random slopes for time and supplement and their interaction) to less complex. The most complex random effect structure that converged was chosen as the final model [@matuschek_balancing_2017]. Plasma glucose, serum c-peptide, training volume, and the strength index were analyzed by multiple time-point log-fold change score comparisons, using a mixed-effects model with baseline values, time and supplement, and the interaction between time and supplement as fixed effects. These data only supported a random intercept per participant. The linear mixed-effects models were fitted with the lmer function from the lme4 package using the lmerTest package to procure p-values (Satterthwaite's method for approximating degrees of freedom) [@bates_fitting_2014], written for R [@Rcore]. Log-transformed values were expressed as fold changes in visualisations. Descriptive data are presented as mean and standard deviation (SD). Inferential statistics are presented as means with 95% confidence intervals and *p*-values unless otherwise stated. *p* \< 0.05 was considered statistically significant.

@@

Glucose ingestion before and after RT sessions did not lead to a higher mean change of total training session volume, with a mean increase of 17% in both the glucose condition (pre: 5262 ± 1799kg, post: 6319 ± 2256kg, *p* > 0.05) and the placebo condition (pre: 5351 ± 1615kg, post: 6438 ± 2092, *p* > 0.05) from session 1 to session 6, respectively. There were no differences in mean macronutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* \> 0.05 for all, raw data and table available at GitHub repository).

@@

Glucose ingestion before and after RT led to increases in plasma glucose levels compared to baseline by 38% immediately before RT (Figure 1B, 0 min), by 31% during RT (Figure 1B, 15 min) and by 32% immediately after RT (Figure 1B, 30 min; all : *p* \< 0.001), with no changes observed in the placebo condition (*p* \> 0.05). Compared to the placebo condition, ingestion of glucose increased plasma glucose levels by 36% immediately before RT (Figure 1B, 0 min), by 27% during RT (Figure 1B, 15 min) and by 28% immediately after RT (Figure 1B, 30 min; all : *p* \< 0.001). Two hours after the RT session, glucose ingestion was associated with 12% lower plasma glucose levels compared to baseline, and 8% lower compared to placebo (Figure 1B, 270min: *p* = `r p.270`).

Glucose ingestion before and after RT led to increases in levels of c-peptide compared to baseline, by 95% immediately before (Figure 1C, 0 min) and 87% after RT (Figure 1C, 30 min; both *p* \< 0.001), with no changes observed with the placebo condition (*p* \> 0.05). Compared to the placebo condition, ingestion of glucose increased levels of c-peptide by 85% immediately before (Figure 1C, 0 min) and 85% after RT (Figure 1C, 30 min; both *p* \< 0.001).

@@

Glucose ingestion before and after RT sessions did not generally improve skeletal muscle recovery throughout the intervention, neither during the intervention in a rested state (Figure 1D, 23hrs after exercise; Post 2RT: *p* = `r pdiff.c2`, Post 4RT: *p* = `r pdiff.c3`), nor acutely after the sixth and final RT-session (30min post 6RT: *p* = `r pdiff.c5`, 2h post 6RT: *p* = `r pdiff.c6`) or in a rested state after the sixth RT session (23h post 6RT: *p* = `r pdiff.c7`). Glucose ingestion before and after RT compared to placebo was associated with a 7% less reduction in strength after five RT sessions (Figure 1D, *p* = `r pdiff.c4`).

Both RT with glucose and placebo led to significantly reduced strength after the fifth session compared to baseline, by 11 and 18% respectively (Figure 1D, Post 5RT: *p* = 0.000). Comparisons of the acute data gathered from after five sessions until and including 23 hours after the sixth session showed an average increase in strength of 5-9% from RT with glucose and placebo 30 minutes after the sixth RT session (Figure 1D, 30min post 6RT: *p* = `r apc2`) and two hours after the sixth RT session (Figure 1D, 2h post 6RT: *p* = `r apc3`). Twenty-three hours after the last (sixth) RT session, strength was unchanged compared to after the fifth RT session (Figure 1D, 23h post 6RT: *p* = `r apc4`).

@@

The five-session-RT intervention led to increases in total RNA and rRNA per unit muscle weight both in the glucose and the placebo condition, on average by \~20-27% and \~25-57%, respectively (Figure 2). However, RT with glucose did not induce increased accumulation of total RNA (Figure 2A, mean difference `r total_rna_stats`) or rRNA (Figure 2B; 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`) compared to RT with placebo.

@@

The datasets generated during and analysed during the current study are available in the "ribose-paper" repository; <https://github.com/Kristianlian/ribose-paper>

@@

Supplementary material can be found in the readme file of the "ribose-paper" repository, available here; <https://github.com/Kristianlian/ribose-paper>

\pagebreak

## References

::: {#refs}
:::
